\documentclass{article}

\usepackage{fullpage}
\usepackage{parskip}

\begin{document}
\title{Program Synthesis for DNA Strand Displacement}
\author{Kendall Stewart, Martin Kellogg, Anthony Ha}

\maketitle

DNA computing seeks to leverage the chemical properties
of DNA molecules as a computational substrate.
From a medical perspective, DNA computing can enable the
construction of complex devices at a cellular scale for more precise
diagnostic tests or drug delivery. From a computer architecture perspective,
DNA is capable of performing massively parallel computations with very low
power consumption.

The feasibility of DNA computing was first demonstrated by encoding
the Hamiltonian Path Problem into DNA~\cite{adelman}.
This encoding was constructed by
hand, but recent work has advanced DNA strand displacement
reactions~\cite{strands} as a primitive for DNA computation. In particular,
strand displacement reactions can be used to implement boolean logic gates,
allowing them to simulate traditional silicon-based computers.

The boolean completeness of strand displacement reactions allows them to be
a conceptual ``machine code" for DNA computing. An assembler-analogue
can convert boolean gates into strand displacement reactions, allowing
programs for DNA-based machines to be written using traditional languages.
However, translating from boolean gates to stand displacement reactions
results in a strand displacement system whose complexity is exponential in the
size of the problem (as expressed in boolean gates). This exponential blowup
prevents this technique from scaling to large systems.

However, strand displacement reactions are capable of more than just boolean
logic. An operation could be implemented with a set of fewer, more
complex strand displacement gates, but reasoning about the behavior of more
complex reactions is difficult for humans.
We hypothesize that a program synthesis tool
equipped with appropriate semantics for strand displacement reactions could
``discover" gates that would be difficult for a human to find.

A formal semantics for strand displacement reactions is provided in the
description of DSD~\cite{dsd}, a programming language for specifying and
simulating DNA strand displacement circuits. Given a set of initial strand
species (reactants), DSD applies reduction rules to infer the intermediate and
final strand species (products), and creates a directed reaction graph with
edges between reactants and products, weighted by the rate of each reaction.
This graph, along with ordinary differential equations derived from the rates,
are used to simulate the strand displacement system.

Ignoring rates, the structure of the unweighted reaction graph encodes
conditions that are necessary (but not sufficient) for some product to have a
particular meaning. For instance, if all paths leading to a product include both
initial species $A$ and initial species $B$, the product could potentially
represent $A \land B$.

As a first pass at applying program synthesis to DNA strand displacement, we
propose to implement an interpreter for the semantics of DSD in Rosette, and
then construct predicates that can reason about the structure of the resulting
graph (without reaction rates). We will then see whether the solver can satisfy
queries such as ``given reactant $A$ and reactant $B$, find a system of gates
where the product represents $A \land B$" (where this meaning is derived from
the graph structure, as described above).

If this is successful (and we have additional time), we will move on to more
complicated queries such as ``given reactants representing number $X$, and
reactants representing the number $Y$, find a system where the output represents
$X + Y$".  It is our hypothesis that the resulting systems may be smaller (in
the number of gates or number of reactions) than systems that are compiled from
known implementations of boolean circuits, because the solver will be able
to take advantage of the special properties of strand displacement reactions.
This hypothesis can be tested by
encoding it as additional constraints for the synthesis problem.

To summarize, our proposed timeline of work for the rest of the term is as
follows:
\begin{enumerate}
\item
Write and test interpreter that implements DSD semantics in Rosette (2
weeks)

\item
Write and test predicates about graph structure (1 week)

\item
Synthesize simple systems implementing boolean circuits (1 week)

\item
Synthesize more complex systems (if time allows)

\item
Prepare final report and presentation (1 week)
\end{enumerate}

\begin{thebibliography}{9}

\bibitem{adelman}
L. M. Adleman,
“Molecular computation of solutions to combinatorial problems,”
Science, vol. 266, no. 5187, pp. 1021–1024, Nov. 1994.

\bibitem{strands}
D. Y. Zhang and G. Seelig,
“Dynamic DNA nanotechnology using strand-displacement reactions,”
Nature Chemistry, vol. 3, no. 2, pp. 103–113, Feb. 2011.

\bibitem{dsd}
M. R. Lakin, S. Youssef, L. Cardelli, and A. Phillips,
“Abstractions for DNA circuit design,”
Journal of The Royal Society Interface, vol. 9, no. 68, Jul. 2011.

\end{thebibliography}

\end{document}
